---
title: "MEDICAL INSURANCE DATA ANALYSIS PROJECT"
author: "PURITY ELIZABETH-DATA ANALYST"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setting a Working directory
```{r}
setwd("C:/Users/BOBBY/Desktop/R Programming")
```

# Data Importation
```{r}
gss <- read.csv("insurance_data.csv")
head(gss)
```

```{r}
tail(gss)
```

```{r}
str(gss)
```

# Check for missing values
```{r}
missing_value <- sum(is.na(gss))
print(missing_value)
```

```{r}
sum(duplicated(gss))
```

```{r}
gss <-gss[!duplicated(gss),]
```

# Detecting Outliers
```{r}
boxplot(gss$age,main = "age")
boxplot(gss$bmi,main = "bmi")
boxplot(gss$children,main = "children")
boxplot(gss$charges,main = "charges")
```

# Handling outliers using 1.5 IQR Rule
```{r}
Q1 <- quantile(gss$bmi, 0.25)
Q3 <- quantile(gss$bmi, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR
d_cleaned <- gss[gss$bmi >= lower_bound & gss$bmi <= upper_bound, ]
```

```{r}
boxplot(d_cleaned$bmi,main = "bmi")
```

```{r}
Q1 <- quantile(gss$charges, 0.25)
Q3 <- quantile(gss$charges, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR
d_cleaned <- gss[gss$charges >= lower_bound & gss$charges <= upper_bound, ]
```

```{r}
boxplot(d_cleaned$charges,main = "charges")
```

```{r}
Q1 <- quantile(d_cleaned$charges, 0.25)
Q3 <- quantile(d_cleaned$charges, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR
g_cleaned <- d_cleaned[d_cleaned$charges >= lower_bound & d_cleaned$charges <= upper_bound, ]
```

```{r}
boxplot(g_cleaned$charges,main = "charges")
```

```{r}
Q1 <- quantile(g_cleaned$charges, 0.25)
Q3 <- quantile(g_cleaned$charges, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR
s_cleaned <- g_cleaned[g_cleaned$charges >= lower_bound & g_cleaned$charges <= upper_bound, ]
```

```{r}
boxplot(s_cleaned$charges,main = "charges")
```

```{r}
Q1 <- quantile(s_cleaned$charges, 0.25)
Q3 <- quantile(s_cleaned$charges, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR
t_cleaned <- s_cleaned[s_cleaned$charges >= lower_bound & s_cleaned$charges <= upper_bound, ]
```

```{r}
boxplot(t_cleaned$charges,main = "charges")
```

```{r}
Q1 <- quantile(t_cleaned$charges, 0.25)
Q3 <- quantile(t_cleaned$charges, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR
r_cleaned <- t_cleaned[t_cleaned$charges >= lower_bound & t_cleaned$charges <= upper_bound, ]
```

```{r}
boxplot(r_cleaned$charges,main = "charges")
```

# Summary Statistics
```{r}
summary(r_cleaned)
```
 

# EDA
```{r}
hist(r_cleaned$age,
     main = "Age Distribution",
     xlab = "Age in Years",
     col ="purple" )
```

> Age has a uniform distribution

```{r}
hist(r_cleaned$bmi,
     main = "BMI Distribution",
     xlab = "Body Mass Index",
     col ="purple" )
```

> BMI distribution has a normal distribution

```{r}
hist(r_cleaned$charges,
     main = "Charges Distribution",
     xlab = "Charges in Kshs",
     col ="green" )
```

> Charges distribution has a positively skewed distribution

```{r}
hist(r_cleaned$children,
     main = "Children Distribution",
     xlab = "Children",
     col ="blue" )
```
> The above histogram  shows a positively skewed distribution  

```{r}
library(ggplot2)
ggplot(r_cleaned, aes(x = age, y = charges))+
  geom_point(color ="purple")+
  labs(title = "Charges vs Age",x = "Age in Years", y = "Charges in Kshs")+
theme_minimal()
```
> The scatter plot indicates that age is a strong predictor of insurance charges implying as age increases charges also increases
The graph shows a clear positive relationship between age and charges, meaning costs tend to rise as people get older.


# Data Manipulation

```{r}
r_cleaned_filtered <- r_cleaned |>
dplyr::filter(age>25 , smoker=="no" , region=="northwest")
head(r_cleaned_filtered)
```

```{r}
r_cleaned_selected <- r_cleaned |>
  dplyr ::select(sex,charges,age )
head(r_cleaned_selected)
```

```{r}
r_cleaned_mutated <- r_cleaned|>
 dplyr :: mutate(charges_category=ifelse(charges >15000,"High","Low"))
head(r_cleaned_mutated)
```

```{r}
r_cleaned_arranged <- r_cleaned |>
dplyr :: arrange(charges)
head(r_cleaned_arranged,20)
```

```{r}
r_cleaned |>
dplyr :: group_by(sex) |>
 dplyr :: summarize(mean_charges=mean(charges,na.rm=TRUE))
```

```{r}
r_cleaned |>
dplyr :: group_by(smoker) |>
 dplyr :: summarize(mean_charges=mean(charges,na.rm=TRUE))

```

```{r}
r_cleaned |>
dplyr :: group_by(region) |>
 dplyr :: summarize(mean_bmi=mean(bmi,na.rm=TRUE))
```

```{r}
r_cleaned_processed <- r_cleaned|>
  dplyr::filter(age > 25) |>
  dplyr::select(region ,charges, age) |>
  dplyr::mutate(age_group = ifelse(age > 25, "Adults", "Minor")) |>
  dplyr::arrange(charges,10)
head(r_cleaned_processed)
```

# Data Visualization
## Scatter Plot
```{r}
library(ggplot2)
r_cleaned |> ggplot (aes(x=age,y=bmi,color=smoker)) +
                geom_point(size = 2, alpha=0.8)+
  labs(x="Age in Years",
       y="Body Mass Index",
       title="Scatter plot showing relationship betweeen Age and BMI",
       caption="Source:Insurance Dataset,2026")+
  theme_bw()
```
BMI values are spread across all age groups, suggesting there is no strong relationship between age and BMI. Most individuals have BMI values between 20 and 40, and non-smokers appear to make up the majority of the data points. Smokers and non-smokers are distributed similarly across different ages and BMI ranges.


## Bar  Chart
```{r}
r_cleaned |>
  ggplot(aes(x=smoker,fill=smoker))+
  geom_bar(color="black")+
  labs(x="Smoker",
       y="No of Respondents",
       title="A Bar Chart showing Smoker distribution",
       caption="Source: Insurance Dataset,2026")+
  theme_bw()
```
Non-smokers make the largest number in the data set

## Pie Chart
```{r}
library(tidyverse)

r_cleaned |>
  group_by(region) |>
  summarise(total_charges = sum(charges, na.rm = TRUE)) |>
  mutate(
    pct = total_charges / sum(total_charges),
    label = paste0(region, "\n", scales::percent(pct, accuracy = 0.1))
  ) |>
  ggplot(aes(x = "", y = total_charges, fill = region)) +
  geom_col(width = 1) +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4) +
  coord_polar(theta = "y") +
  labs(
    title = "Total Charges by Region",
    fill = "Region"
  ) +
  theme_void()
```
The northeast region accounts for the highest share of total charges at 27.0%. The northwest and southeast regions contribute equally at 24.4% each. The southwest region has the lowest share of total charges at 24.2%.


## Boxplot
```{r}
library(ggplot2)
r_cleaned |>
  ggplot(aes(x=region,y= charges,fill=sex))+
  geom_boxplot(color="black")+
  labs(x="Region distribution",
       y="Charges in Kshs",
       title="A Boxplot of Charges by Region",
       caption="Source: Insurance Dataset,2026")+
  theme_bw()
```
The graph shows that insurance charges vary across regions and between males and females.The southeast region has the highest median charges for both sexes, while the northeast and northwest regions have relatively lower and similar distributions. Male charges tend to be slightly higher than female charges in most regions, and there are a few outliers indicating some individuals with  high insurance Charges.

```{r}
library(plotly)
p=r_cleaned |> 
  ggplot(aes(x= region, y = charges, fill = smoker)) +
  geom_boxplot() + 
  labs(x = "Region distribution",
       y = " Charges in Kshs",
       title =  "Boxplot showing  Charges Distribution by Region and Smoker",
       caption = "Source: Insurance Dataset, 2026") +
       theme_classic() + 
  theme(legend.position = "bottom")

ggplotly(p)
```

The boxplot shows that Insurance charges vary across all regions, but the  distribution is quite similar. In every region, smokers tend to have much higher charges than non-smokers, as seen by the higher median values and wider spread in the upper box plots. There are also several outliers, indicating some individuals have extreme high medical expenses. Overall, smoking appears to have a stronger impact on charges than the region itself.


# Hypothesis Testing
## Two Sample t_test
HO:Mean insurance charges of Male is equal to mean insurance charges of Female
H1:Mean insurance charges of Male is not equal to mean insurance charges of Female
```{r}
result <- t.test(charges~sex, data=r_cleaned)
result
```
From the results above where p-value is greater than 0.05,i  fail to reject  Ho and conclude that  the Mean Insurance charges of Male and Female is equal to each other 

HO:There is no significant difference in mean Body Mass Index between Smokers and non-smokers
H1:There is  significant difference in mean Body Mass Index between Smokers and non-smokers
```{r}
result <- t.test(bmi~smoker, data=r_cleaned)
result
```
The t-test results indicate a significant difference in mean Body Mass Index between smokers and non-smokers. The p-value is less than 0.05,indicating that the  difference is statistically significant.

## Chi Square Test
HO:There is no significant association  between smoking status gender
H1:There is  significant association  between smoking status gender
```{r}
library(gtsummary)
tbl_summary (r_cleaned[c("sex","smoker")],
             by="smoker")%>%
  add_ci()%>%
  add_p(
    test=list(
      all_categorical()~"chisq.test"
     
    )
  ) 
 
```
Having my p-value>0.05 i fail to reject H0 at 5% Level of Significance and conclude that there is no significant association between smoking status and gender 

# Correlation Analysis
## Heatmap
```{r}
library(corrplot)
r_cleaned |>
  select(age,bmi,charges) |>
  cor(use="complete.obs") |>
  corrplot(method = "color",
           addCoef.col = "black",tl.col = "black")
        
```
The heatmap shows that age has a moderate positive relationship with charges , meaning insurance costs tend to rise as a person gets older. In contrast, there is a very weak positive relationship between age and BMI , and a negative relationship between BMI and charges ,indicating that BMI alone does not strongly correlate with changes in charges or age in this dataset.

# Regression Analysis
## Simple Linear Regression
HO:Age has no significant effect on charges
H1: Age has a significant effect on charges
```{r}
model <- lm(charges ~ age, data = r_cleaned)
summary(model)
```
From the results my p-value is less than 0.05 at 5% level of significance,therefore conclude that age has a significant effect on charges.As one unit of age increases charges by 203.25 units.

## Multiple Linear Regression
HO:Age,BMI,Children,Sex,Smoker do not have a significant effect on charges
H1:At least one of the variables has a significant effect on charges
```{r}
model <- lm(charges ~ age+bmi+children+sex+smoker, data = r_cleaned)
summary(model)
```
The model expresses charges by 69.49 percentage.From the results  p-value of Age  is less than 0.05 at 5% level of significance,therefore conclude that age has a significant effect on charges.As one unit of age increases charges by 203.25 units.As Body Mass Index increases by a unit it increases charges by 15.965 units.Since p-value is greater than 0.05,it indicates that body mass index do not have a significant effect on charges. For each additional child ,charges increases by 409.989 units.Since p-value is less than 0.05,it indicates that children have a significant effect on charges.Male have  336.307 unit charges lower than females. Since p-value is greater than 0.05,it indicates that male do not have a significant effect on charges.Smokers pay charges by 12823.326 units more than non-smokers. Since p-value is less than 0.05,it indicates that male  have a significant effect on charges.Thus,reject HO and conclude that atl east one of the variables has a significant effect on charges


# SPECIAL THANKS TO CENTER FOR DATA ANALYTICS AND MODELLING-CHUKA UNIVERSITY.
# CERTIFICATION COURSE:INTRODUCTION TO R FOR DATA SCIENCE.






